### 2022-12-09 introduction/introduction.tex
Converted protein outline into text and tweaked a few phrases elsewhere

Life is a physical phenomenon. Despite the complexity of living things, their processes are governed by the same physical laws that describe how the planets revolve around the sun and the probability that an atom of uranium decays from one moment to another. However, as many systems are too complex to describe with physical models and equations, scientists simplify them into levels of abstraction that are more useful. (This is the origin of the common observation that biology is applied chemistry and chemistry is applied physics.) For example, Punnett squares facilitate the prediction of genotypes and phenotypes by distilling the complexities and nuances of diverse reproductive systems into a set of simple rules. However, since life spans a scale from single cells to entire ecosystems, biology likely employs more layers of abstraction than any other scientific discipline. One of the most powerful and widely used frameworks within the life sciences is biochemistry, which characterizes biological processes in terms of their component molecules and chemical reactions. A specific focus is four classes of macromolecules called nucleic acids, carbohydrates, lipids, and proteins, all of which are unique to biological systems. Though not all biological molecules are macromolecules\footnote{Macromolecule is a loose term applied to molecules with high molecular mass. In practice, it usually refers to one of the classes listed above, among a few other prominent non-biological examples.} and not all biological macromolecules fit neatly in one of these four categories, much of life at the molecular level is understood in terms of their structure and function. Each of the four has a characteristic role. Nucleic acids, i.e., DNA and RNA, are responsible for information storage and transfer. Carbohydrates primarily store energy but can also act as structural components of cells. Lipids are a diverse class of oily molecules which are components of cell membranes, store energy, and transmit signals. Proteins have a range of functions, including catalyzing reactions, transmitting signals, transporting materials, and providing structure. Many of these overlap with the functions of the other macromolecule classes because proteins are involved in virtually every biological process. However, unlike the others, which are often passive participants, proteins are highly active and dynamic. They respond to signals, change shape, and often use the other macromolecules as substrates in their activities. Proteins are essentially the molecular machines that carry out life's functions.

Some examples will illustrate the central role of proteins more clearly.
These examples should be accessible to a wide audience, so nothing too abstract. Ideally should relate to human physiology and disease.
    1. Hemoglobin and sickle cell
        A. Introduce red blood cells transport oxygen
        B. Hemoglobin is protein that binds oxygen.
            i. Physically interacts with a molecule of oxygen
        C. Sickle cell is disease caused by mutation in hemoglobin protein.
            i. Cells produce hemoglobin with small mistake which causes them to clump into chains
            ii. This distorts the shape of red blood cells into a sickle
            iii. Red blood cells do not flow as easily through blood vessels causing pain and oxygen deprivation
    2. Photopsins and color blindness
        A. Photopsins are light sensitive proteins in cells in the retina at the back of the eye
        B. Three different versions are sensitive to red, green, and blue light
        C. Color blindness is the result of cells missing one of these proteins
    3. Lactase and lactose intolerance
        A. Lactose is a sugar found in milk that requires specific protein, lactase, to digest
        B. Many humans who can digest milk products in childhood, lose this ability in adulthood suffer symptoms such as bloating and diarrhea after consuming dairy products
        C. They stop producing lactase, so the undigested sugars are broken down by bacteria in the colon

Despite performing this diverse range of functions, all proteins are made from of a set of 20 simple building blocks called amino acids\footnote{The term amino acid encompasses any compound that contains an amino and carboxyl group. However, proteins are only synthesized from the twenty ``canonical'' amino acids. Another two (selenocysteine and pyrrolysine) are incorporated via a distinct mechanism under rare circumstances and are therefore considered non-standard.}. Though each amino acid is chemically unique, they share a common backbone composed of two distinct and complementary receptor and donor sites for chemical bonds. Thus, in a protein the amino acids are bonded in a linear chain like beads on a string. However, once synthesized, proteins are not tidy rod-shaped molecules. Instead, the chain loops and weaves between itself creating a characteristic three-dimensional structure in a process called folding. These structures, which are highly stable and characteristic of each protein, are a result of the interactions between the amino acids in the chain and the surrounding medium, which is typically water. Because each amino acid has unique geometric and chemical properties that influence the energetics of these interactions, a protein's three-dimensional structure is encoded by the sequence of amino acids that compose it. A protein's function is in turn a direct result of its structure. For example, the structure of hemoglobin precisely positions its amino acids and a helper molecule called a heme group to create a pocket that can stably but reversibly bind oxygen. This allows hemoglobin to carry oxygen throughout the body until it is delivered to its destination. However, people affected by sickle cell disease have a defect, i.e., a mutation, in the sequence of their hemoglobin proteins which causes it to malfunction. Frequently this mutation is a single change where the sixth amino acid in the sequence, glutamate, is substituted for a valine. This creates a sticky patch on the surface of hemoglobin, and under low-oxygen conditions normal hemoglobin changes shape to expose a sticky patch on its surface as well. The two patches are complementary, which allows hemoglobin proteins to clump together into long, fibrous strands. These strands distort the shape of red cells, giving them their characteristic sickle shape.

Because sequence dictates structure which in turn dictates function, understanding these relationships in proteins is essential for understanding biological systems as a whole. While biologists are generally interested in function since it is the most directly associated with the biological role\footnote{Function and biological role are frequently, though not always, exact synonyms since biological role is often defined more broadly and loosely than function. For example, the strict function of hemoglobin is to bind oxygen, an activity it shares with a related protein myoglobin. However, the two have different roles. Hemoglobin is found in red blood cells where it primarily acts a carrier. In contrast, myoglobin is found in muscle cells, where it stores oxygen until needed.} of a protein, function and biological role are not always easily identified or measured. Thus, determining a protein's structure is frequently the first step of detailed study of a protein's function. Though in recent years researchers have developed powerful computational tools that can accurately predict structure from sequence alone, historically structures were determined experimentally, and experimental methods remain the gold standard today. While many methods can reveal information about the structure of a protein, the most powerful techniques, X-ray crystallography and NMR spectroscopy, can map the spatial coordinates of every atom in a protein. However, this resolution requires extremely pure samples of the protein of interest. Since proteins are only produced by living systems, preparations begin with a complex mixture consisting of cells or tissue, and the protein of interest is isolated with series of extraction and purification steps. Some proteins are only produced in small amounts or degrade easily, so each step may require substantial optimization to achieve a sufficient yield. When structural techniques were first developed in the late 1950s, they were so time-consuming that a graduate student could dedicate an entire PhDs to solving a single protein structure. Many developments have substantially accelerated the process, but it remains a labor-intensive technique which may require several months of effort. However, the result is a powerful map that scientists use to suggest hypotheses and interpret data.

>>>>

Life is a physical phenomenon. Despite the complexity of living things, their processes are governed by the same physical laws that describe planets' motion around the sun and the decay of uranium atoms into lighter elements. However, as many systems are too complex to describe with physical models and equations, scientists simplify them into levels of abstraction that are more useful. (This is the origin of the common observation that biology is applied chemistry and chemistry is applied physics.) For example, Punnett squares facilitate the prediction of genotypes and phenotypes by distilling the complexities and nuances of diverse reproductive systems into a set of simple rules. However, since life spans a scale from single cells to entire ecosystems, biology likely employs more layers of abstraction than any other scientific discipline. One of the most powerful and widely used frameworks within the life sciences is biochemistry, which characterizes biological processes in terms of their component molecules and chemical reactions. A specific focus is four classes of macromolecules called nucleic acids, carbohydrates, lipids, and proteins, all of which are unique to biological systems. Though not all biological molecules are macromolecules\footnote{Macromolecule is a loose term applied to molecules with high molecular mass. In practice, it usually refers to one of the classes listed above, among a few other prominent non-biological examples.} and not all biological macromolecules fit neatly in one of these four categories, much of life at the molecular level is understood in terms of their structure and function. Each of the four has a characteristic role. Nucleic acids, i.e., DNA and RNA, are responsible for information storage and transfer. Carbohydrates primarily store energy but can also act as structural components of cells. Lipids are a diverse class of oily molecules which are components of cell membranes, store energy, and transmit signals. Proteins have a range of functions, including catalyzing reactions, transmitting signals, transporting materials, and providing structure. Many of these overlap with the functions of the other macromolecule classes because proteins are involved in virtually every biological process. However, unlike the others, which are often passive participants, proteins are highly active and dynamic. They respond to signals, change shape, and often use the other macromolecules as substrates in their activities. Proteins are essentially the molecular machines that carry out life's functions.

Some examples will illustrate the central role of proteins more clearly. Blood is a part of the circulatory system which is responsible for transporting nutrients and waste. Though blood is a complex mixture, containing a cocktail of cells, proteins, sugars, gases, and ions dissolved in a medium of water, its primary cellular component is red blood cells. These cells, which give blood its red color, ferry oxygen from lungs throughout the body. While water can dissolve some oxygen, the body requires more oxygen more quickly than is available in the aqueous component of blood alone. Thus, red blood are packed with a special protein called hemoglobin, which binds\footnote{In biology, ``binding'' is a slippery word whose exact meaning can varying greatly depending on the context, but it generally means that a molecule physically interacts with another molecular for an extended period.} oxygen. Each red blood cell contains as many as 270 million molecules of hemoglobin, each of which can carry up to four oxygen molecules. To put this value in perspective, if each molecule of hemoglobin were the size of a tennis ball, the hemoglobin molecules in a single red blood cell would cover a tennis court in over 10 packed layers\footnote{A standard tennis ball is 1.35 in or 0.1125 ft in radius, so the volume of 270 million tennis balls of this size packed as efficiently as possible is given by $2.70 \times 10^6 \frac{4}{3} \pi (0.1125)^3 \frac{3\sqrt{2}}{\pi} = 2.17 \times 10^4 \text{ft^2}$. A singles tennis court is 78 ft by 27 ft or 2106 ft\textsuperscript{2}, so dividing the first by the second yields slightly over 10 layers.}. Because red blood cells are so dense with hemoglobin, any defect in hemoglobin can dramatically impact the structure of the red blood cells themselves. A well-studied example is sickle cell disease where an error in the body's hemoglobin molecules deforms red blood cells into a characteristic sickle shape. This prevents them from easily flowing through blood vessels, resulting in pain and oxygen deprivation.

Whereas hemoglobin is an example of a protein mediating transport, proteins are also involved in transmitting signals and catalyzing chemical reactions. For example, the back of the eye contains a light-sensitive surface called the retina which is composed of photoreceptor cells. These cells respond to light because they produce special proteins called opsins that translate light into chemical and electrical signals which are then interpreted by the brain. Humans, and primates broadly, have three types of opsins that mediate our color vision, which are sensitive to red, green, and blue light, respectively. Color blindness is the result of cells missing one of these proteins, typically either the red or green opsin. In contrast to sickle cell disease, here the condition is caused by a missing rather than a mutated protein. However, in other cases, a protein is not missing or mutated, but instead not produced at the right time and place. For example, lactose is a sugar found in milk that requires a secific protein, lactase, to metabolize properly. Many humans who can digest milk products in childhood, lose this ability in adulthood because they stop producing lactose. As a result, lactose in dairy products passes undigested into the colon where it is broken down by bacteria, causing symptoms such as bloating and diarrhea.

Despite performing this diverse range of functions, all proteins are made from of a set of 20 simple building blocks called amino acids\footnote{The term amino acid encompasses any compound that contains an amino and carboxyl group. However, proteins are only synthesized from the twenty ``canonical'' amino acids. Another two (selenocysteine and pyrrolysine) are incorporated via a distinct mechanism under rare circumstances and are therefore considered non-standard.}. Though each amino acid is chemically unique, they share a common backbone composed of two distinct and complementary receptor and donor sites for chemical bonds. Thus, in a protein the amino acids are bonded in a linear chain like beads on a string. However, once synthesized, proteins are not tidy rod-shaped molecules. Instead, the chain loops and weaves between itself creating a characteristic three-dimensional structure in a process called folding. These structures, which are highly stable and characteristic of each protein, are a result of the interactions between the amino acids in the chain and the surrounding medium, which is typically water. Because each amino acid has unique geometric and chemical properties that influence the energetics of these interactions, a protein's three-dimensional structure is encoded by the sequence of amino acids that compose it. A protein's function is in turn a direct result of its structure. For example, the structure of hemoglobin precisely positions its amino acids and a helper molecule called a heme group to create a pocket that can stably but reversibly bind oxygen. This allows hemoglobin to carry oxygen throughout the body until it is delivered to its destination. However, people affected by sickle cell disease have a mutation in the sequence of their hemoglobin proteins which causes it to malfunction. Frequently this mutation is a single change where the sixth amino acid in the sequence, glutamate, is substituted for a valine. This creates a sticky patch on the surface of hemoglobin, and under low-oxygen conditions normal hemoglobin changes shape to expose a sticky patch on its surface as well. The two patches are complementary, which allows hemoglobin proteins to clump together into long, fibrous strands. These strands distort the shape of red cells, giving them their characteristic sickle shape.

Because sequence dictates structure which in turn dictates function, understanding these relationships in proteins is essential for understanding biological systems as a whole. While biologists are generally interested in function since it is the most directly associated with the biological role\footnote{Function and biological role are frequently, though not always, exact synonyms since biological role is often defined more broadly and loosely than function. For example, the strict function of hemoglobin is to bind oxygen, an activity it shares with a related protein myoglobin. However, the two have different roles. Hemoglobin is found in red blood cells where it primarily acts a carrier. In contrast, myoglobin is found in muscle cells, where it stores oxygen until needed.} of a protein, function and biological role are not always easily identified or measured. Thus, determining a protein's structure is frequently the first step of detailed study of a protein's function. Though in recent years researchers have developed powerful computational tools that can accurately predict structure from sequence alone, historically structures were determined experimentally, and experimental methods remain the gold standard today. While many methods can reveal information about the structure of a protein, the most powerful techniques, X-ray crystallography and NMR spectroscopy, can map the spatial coordinates of every atom in a protein. However, this resolution requires extremely pure samples of the protein of interest. Since proteins are only produced by living systems, preparations begin with a complex mixture consisting of cells or tissue, and the protein of interest is isolated with series of extraction and purification steps. Some proteins are only produced in small amounts or degrade easily, so each step may require substantial optimization to achieve a sufficient yield. When structural techniques were first developed in the late 1950s, they were so time-consuming that a graduate student could dedicate an entire PhD to solving a single protein structure. Many developments have substantially accelerated the process, but it remains a labor-intensive technique which may require several months of effort. However, the result is a powerful map that scientists use to suggest hypotheses and interpret data.
